---
title: "Statistical Inference Course Project - Data Analysis"
author: "Daniel Pont"
date: '2018-09-30'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(cowplot)
set.seed(1)
```

## Synopsis
This report is the second part of the Statistical Inference Course Project. It's a basic inferential data analysis of the ToothGrowth dataset that shows that vitamin C increases tooth growth in guinea pigs. It also shows that for dose < 2 tooth growth is higher with
OJ than with VC.  

## Basic Inferential Data Analysis of the dataset ToothGrowth    

### 1) Exploratory Data Analysis    

```{r data_analysis_plot, message=FALSE, warning=FALSE,fig.width=7, fig.height=3}

ToothGrowth$dose_factor <- as.factor(ToothGrowth$dose)

p_all <- ggplot(ToothGrowth, aes(x=dose_factor, y=len)) + 
    geom_boxplot() +
    ggtitle("Plot of teeth length \n by dose") +
    xlab("Dose") + ylab("Len")

p_by_supp <- ggplot(ToothGrowth, aes(x=dose_factor, y=len, fill=supp))+     geom_boxplot() +
    ggtitle("Plot of teeth length \n by dose and supp") +
    xlab("Dose") + ylab("Len")

plot_grid(p_all, p_by_supp, 
          ncol = 2, nrow = 1)

```

* Tooth growth seems to increase with vitamin C dose
* Tooth growth seems to increase more with OJ than with VC for dose < 2
* Tooth growth doesn't seem different with OJ or VC for dose = 2 

### 2) Summary of the data
```{r}
ToothGrowth %>% group_by(supp,dose) %>%  
    summarize(mean=mean(len), sd=sd(len))
```

### 3) Comparison of tooth growth by supp and dose  

According to 2) "Summary of the data", the variance of each group is different. We assume the groups are not paired. So we can perform t-tests with (paired=FALSE,var.equal=FALSE) for the following couple of datasets :

* len for dose = 2  vs len for dose = 0.5 (all supp)
* len for supp = OJ vs len for supp = VC (dose <2)
* len for supp = OJ vs len for supp = VC (dose =2)  

***

```{r message=FALSE,warning=FALSE}

dose0.5 <- ToothGrowth %>% filter(dose==0.5) %>% select(len)
dose2.0 <- ToothGrowth %>% filter(dose==2) %>% select(len)

t.test(dose2.0,dose0.5,paired=FALSE)

OJ.low.dose <- ToothGrowth %>% filter(supp=="OJ",dose<2) %>% select(len)
VC.low.dose <- ToothGrowth %>% filter(supp=="VC",dose<2) %>% select(len)

t.test(OJ.low.dose,VC.low.dose,paired=FALSE)

OJ.high.dose <- ToothGrowth %>% filter(supp=="OJ",dose==2) %>% select(len)
VC.high.dose <- ToothGrowth %>% filter(supp=="VC",dose==2) %>% select(len)

t.test(OJ.high.dose,VC.high.dose,paired=FALSE)
```

*** 

To sum-up :

| len means difference | 95% Confidence Interval | P-value |
| -------------------- | ----------------------- | ------- | 
| dose=2 - dose=0.5    | [12.8,18.2]             | 4.4e-14 |
| OJ - VC (dose <2)    | [1.88,9.30]             | 0.0042  |
| OJ - VC (dose =2)     | [-3.80,3.64]            | 0.96    |


### 4) Conclusion    
Under the following assumptions :

* the  t-tests with default variance parameter (unequal variance) is relevant
* groups are not paired  (different guinea pigs are used for different supp and dose combinations)

We found that :

* Tooth growth is significantly increased when dose is increased from 0.5 to 2  
(p-value << 0.05 and the 95% confidence interval minimum is strictly positive)
* Tooth growth is significantly increased for dose <2 when OJ is used instead of VC 
(same reason than above)
* For dose =2, there's no significant difference in tooth growth between OJ and VC
( p-value >> 0.05, confidence interval centered on 0)
